Concurrent Programming with CXL

### 2020-06-07 book/paper.tex
Mesa monitors

\texttt{wait} releases the monitor mutex and blocks while trying
to procure the condition variable (which is a semaphore that is 0 when
\texttt{wait} is invoked).
\texttt{signal} vacates the condition variable, passing the baton to
a process trying to procure it, and then procures the monitor mutex.
The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
\texttt{V} operations alternate as prescribed by the SBS technique.
Our implementations of reader/writer locks can be easily changed to
use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
\texttt{V}.

In the late 70s, Xerox PARC developed a new programming language called
Mesa.  Mesa introduced various important concepts to programming languages,
including software exceptions and incremental compilation.  Mesa also
incorporated a version of monitors.
However, there are some subtle but important differences with Hoare-style
monitors that make Mesa-style monitors quite unlike split binary semaphores.

>>>>

\texttt{wait} releases the monitor lock and blocks while trying
to procure the condition variable (which is a semaphore that is 0 when
\texttt{wait} is invoked).
\texttt{signal} vacates the condition variable, passing the baton to
a process trying to procure it, and then procures the monitor lock.
The use of \texttt{wait} and \texttt{signal} ensures that \texttt{P} and
\texttt{V} operations alternate as prescribed by the SBS technique.
Our implementations of reader/writer locks can be easily changed to
use \texttt{wait} and \texttt{signal} instead of using \texttt{P}
\texttt{V}.

In the late 70s, Xerox PARC developed a new programming language called
Mesa~\cite{mesa}.
Mesa introduced various important concepts to programming languages,
including software exceptions and incremental compilation.  Mesa also
incorporated a version of monitors.
However, there are some subtle but important differences with Hoare
monitors that make Mesa monitors quite unlike split binary semaphores.

As in Hoare monitors, there is a lock associated with each monitor,
and the lock is acquired upon entry to a method and released upon exit.
Mesa monitors also have condition variable that a process can wait on.
Like in Hoare monitors, the \texttt{wait} operation release the monitor
lock.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
When a process $p$ invokes \texttt{notify}, it does not immediately pass 
he baton to some process $q$ requiring the baton.  Instead, $p$ holds onto
the baton until it leaves the monitor.  At that point $q$ gets the baton,
assuming there even was such a process $q$.

\begin{figure}
\begin{center}
\includegraphics[width=6in]{figures/monitor.jpeg}
\end{center}
\caption{High-level depictions of Hoare and Mesa monitors.  The hall is
where processes wait to enter the bathroom (the critical section).  The
bedrooms illustrate two condition variables.  The circles are processes.}
\label{fig:pc}
\end{figure}

Figure~\ref{fig:bathroom} illustrates the difference with a drawing.
Here a bathroom represents the critical section, allowing only one
process at a time say.  The bedrooms represent condition variables.
There is some condition associated with each bedroom.
In the hall are processes waiting to enter the critical section.

On the left is a Hoare monitor.  When a process leaves the monitor
(let's call that process $p$), it must be in the bathroom.
It can choose to either let a process
in one of the bedrooms or a process in the hall, assuming there are any.
To make the choice, $p$ checks for each bedroom if its condition holds
and if there are processes in the bedroom.
If there are such bedrooms, $p$ selects one and let's that process into
the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
the bathroom, the condition that $q$ was waiting for still holds.

On the right is a Mesa monitor.  The only way into the bathroom is through
the hall.  When process $p$ leaves the bathroom,
if also checks each of the bedrooms to see if there is a process $q$ that
can run.  But instead of letting $q$ go straight into the bathroom, $q$
goes into the hall, joining any other processes that are also waiting to
enter the bathroom.  Finally, when $p$ has left the bathroom, one of the
processes in the hall is let into the bathroom.  But that may not be $q$.
Assuming every process eventually leaves the bathroom and the system is
fair, eventually $q$ will enter the bathroom.  However, it is no longer
guaranteed that its condition holds.  Therefore, the first thing $q$ must
do is check to see if the condition holds.  If not, it should go back
into the bedroom corresponding to the condition.

It turns out that with the Mesa monitor there is another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Note that called \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  Mesa also has a function called
\texttt{notifyAll} (aka \texttt{broadcast)}) that, when applied to
a bedroom, let's all processes in that bedroom into the hall.

While Hoare monitors are conceptually simpler, the so-called
``Mesa monitor semantics'' have become more popular, adopted by all major
programming languages including C, Java, and Python.

\chapter{Mesa Condition Variables in CXL}

\begin{figure}
\begin{code}
\begin{verbatim}
def Condition(lk):
    result = dict{ .lock: lk, .waiters: {} };
;
def wait(c):
    let lk = (^c).lock, blocked = True:
        (^c).waiters = (^c).waiters + { nametag() };
        ^lk = False;
        while blocked:
            atomic:
                if (not (^lk)) and (not ((nametag()) in (^c).waiters)):
                    ^lk = True;
                    blocked = False;
                ;
            ;
        ;
    ;
;
def notify(c):
    let waiters = (^c).waiters:
        if waiters != {}:
            (^c).waiters = waiters - { choose(waiters) };
        ;
    ;
;
def notifyAll(c):
    (^c).waiters = {};
;
\end{verbatim}
\end{code}
\caption{Implementation of condition variables in the \texttt{synch} module.}
\label{fig:cv}
\end{figure}
